---
title: "Precision of binomial overdispersion estimate (simulation)"
author: "Thomas Cornulier"
date: "2024-10-24"
output: 
  html_document:
    self_contained: yes
---

```{r setup, include= FALSE}
knitr::opts_chunk$set(cache= TRUE, out.width = "100%")
```

```{r}
# Print R version
R.version.string
```

# Aims

Here we'll be looking at the effect of the number of trials (binomial denominator) on the ability to estimate binomial overdispersion. A motivating example could be the unexplained variability due to analytical variability across samples, in an eDNA metabarcoding study.

# Simulation functions

```{r}
library(mgcv)
library(parallel)
library(ggplot2)
```

## Data simulation function

Assumes binomial data with:
`nobs`= number of observations, 
`trials`= number of trials per observation,
`p`= probability of success per trial,
`sd`= standard deviation of the between-sample normal error (on the logit scale).

```{r}
r_overbin<- function(nobs, trials, p, sd) {
  # Convert p to logit scale
  logit_p<- log(p / (1 - p))
  # Add Gaussian error
  logit_p_heterog<- rnorm(nobs, mean= logit_p, sd= sd)
  # Convert back to probability scale
  p_adj<- exp(logit_p_heterog) / (1 + exp(logit_p_heterog))
  
  # Simulate binomial data
  return(rbinom(nobs, size= trials, prob= p_adj))
}
```

Test it

```{r}
r_overbin(nobs= 10, trials= 30, p= 0.4, sd= 0.3)
```

(OK)

## Model fit simulation function

```{r}
sim_overbin<- function(nobs, ntrials, p, sd, nsim) {
  dispersion_params<- numeric(nsim)
  obsID<- factor(1:nobs)

  # simulate
  for(i in 1:nsim) {
    simulated_data<- r_overbin(nobs, ntrials, p, sd)
    dat<- data.frame(success= simulated_data, ntrial= ntrials, obsID= obsID)
    fit<- gamm(cbind(success, ntrial - success) ~ 1, random= list(obsID= ~ 1), family= binomial, verbosePQL= F, data= dat)
    dispersion_params[i]<- as.numeric(VarCorr(fit$lme)[1, 2])
    # summary(fit)$dispersion
    cat(".")
  }
  
  return(quantile(dispersion_params, c(0.1, 0.5, 0.9)))
}
```

Test it

```{r}
sim_overbin(nobs= 10, ntrials= 30, p= 0.4, sd= 0.3, nsim= 3)
```

(OK)

# Simulation experiment

## Simulation parameters

Assume:

* sample sizes of either 50 or 500,
* number of trials varying from 2 to 30,
* probability of success p= 0.3,
* overdispersion SD either 0.2 or 0.4, on the logit scale.

We'll run 10000 simulations of each parameter combination for the smaller sample sizes and 1000 for the larger sample sizes.

```{r}
trial_seq<- c(2:5, seq(7, 15, by= 2), 20, 25, 30)
sim_array<- rbind(
	expand.grid(nobs= 50, ntrials= trial_seq, p= 0.3, sd= c(0.2, 0.4), nsim= 10),
	expand.grid(nobs= 500, ntrials= trial_seq, p= 0.3, sd= c(0.2, 0.4), nsim= 10)
	)
```

Visualizing overdispersion values of 0.2 and 0.4 (around p= 0.5):

```{r, out.height = "60%", fig.cap= "Histograms of normal deviates around 0 (on the logit scale), back-transformed onto the probability scale. Thick vertical line shows mean probability, p= 0.5"}
par(mfrow= c(1, 2))
hist(plogis(rnorm(1e6, 0, 0.2)), breaks= (0:100)/100, xlim= c(0, 1), main= "SD= 0.2", freq= F)
abline(v= 0.5, lwd= 2)
hist(plogis(rnorm(1e6, 0, 0.4)), breaks= (0:100)/100, xlim= c(0, 1), main= "SD= 0.4", freq= F)
abline(v= 0.5, lwd= 2)
```

## Results

```{r, eval= T}
# Setup the cluster
num_cores<- 7
cl <- makeCluster(num_cores)
# Export the wrapper function and my_function to the cluster
clusterExport(cl, c("r_overbin", "sim_overbin"))
# run simulation in parallel
tmp<- parApply(cl, sim_array, 1, function(x) {
  require(mgcv)
  sim_overbin(nobs= x[1], ntrials= x[2], p= x[3], sd= x[4], nsim= x[5])
})

stopCluster(cl)

result<- cbind(sim_array, t(tmp))
result$'True SD'<- factor(result$sd)
result$'N obs'<- factor(result$nobs)

save(results, "result.RData")
load("result.RData")

```

```{r, fig.cap= "True and estimated overdispersion SD (dashed and plain lines), with 80% simulation envelopes."}

ggplot(result, aes(x= ntrials, y= `50%`, col= `True SD`, fill= `True SD`)) +
  geom_ribbon(aes(ymin= `10%`, ymax= `90%`), alpha= 0.2) +
  geom_line(aes(y= sd), linetype= "dashed", linewidth= 1.2) +
  geom_line(linewidth= 1.5) +
  facet_wrap(vars(`True SD`, `N obs`), labeller = "label_both") +
  labs(x= "Number of trials", y= "Estimated overdispersion (SD)")

```

* True overdispersion was underestimated in all situations except when both sample size and number of trials were large.
* Minimum number of trials required to get a "reasonable" on-average estimate was quite high, e.g. at least 5 and probably 10-15 for for smaller sample sizes.
* Considerable uncertainty at small sample sizes (e.g., 50).
* Smaller overdispersion harder to estimate precisely.


## In case wanting to run simulations not in parallel

```{r}
# tmp<- apply(sim_array, 1, function(x){
#   sim_overbin(nobs= x[1], ntrials= x[2], p= x[3], sd= x[4], nsim= x[5])
# })
# result<- cbind(sim_array, t(tmp))
# result$'True SD'<- factor(result$sd)

# ggplot(result, aes(x= ntrials, y= `50%`, col= `True SD`, fill= `True SD`)) +
#   geom_ribbon(aes(ymin= `2.5%`, ymax= `97.5%`), alpha= 0.2) +
#   geom_line(aes(y= sd), linetype= "dashed", linewidth= 1.2) +
#   geom_line(linewidth= 1.5) +
#   facet_wrap(~ `True SD`) +
#   labs(x= "Number of trials", y= "Estimated overdispersion (SD)")
```
